### 2022-12-01 intro.Rnw
nicer toc
suggestions from Charlotte

\chapter*{Introduction}
\phantomsection
\addcontentsline{toc}{chapter}{Introduction}

@@

How can we know if a finding from a scientific study is really true? For
example, how can we know if the protective effect of a vaccine found in a study
is real? The answer to such questions is of high importance to scientists and
decision makers. However, because any study result comes with uncertainty we can
never know for sure.

@@

Despite their tremendous importance, the traditional academic system has made it
unattractive for researchers to conduct replication studies; the currencies of
science -- publications, citations, and grant money -- are typically easier to
acquire by conducting novel research. This is because until relatively recently,
top journals in many fields refused to publish replication studies as they were
not considered ``innovative'' enough. To build a successful career, researchers
thus often had no other choice than to concentrate their efforts on producing
novel and eye-catching research results.

@@

\subsection{The statistics of replication studies}
Despite the increased interest in replication studies, the research community
has not yet agreed on one important question: When is a replication study
considered successful? For this reason, replication researchers typically report
the results from different methods for assessing replication success. For
example, the \citet{Opensc2015} states ``[t]here is no single standard for
evaluating replication success. We evaluated [replicability] using significance
and P values, effect sizes, subjective assessments of replication teams, and
meta-analyses of effect sizes'' (p. 11). In the following, I will give an
overview about these and other methods which have been used in practice.

@@

Most methods for analyzing replication studies can be formulated in terms of
study-level summary statistics as shown in Table~\ref{tab:meta}. % Effect
% estimates, standard error, confidence intervals, and $p$-values
All of these are routinely reported in researcher articles, and if one of them
is missing they typically can, under some assumptions, be calculated from the
others. Using summary statistics is also often the only possible way for
conducting the replication success analysis as the raw data from the original
study may not be available to the replicators. The most important statistic is
the effect estimate $\that$. It provides an estimate of the underlying effect
size $\theta$ which quantifies the effect or association of an
intervention/exposure with an outcome variable. Typical effect sizes are mean
differences and correlations (for continuous outcomes), odds ratios (for binary
outcomes), or hazard ratios (for time to event outcomes). Depending on the type
of effect size, a transformation might be required so that the assumption of
approximately normally distributed effect estimates around the unknown effect
size $\theta$ (for large enough sample sizes) is justified. This could be, for
instance, the Fisher $z$-transformation for correlations or the
log-transformation for odds/hazard ratios \citep[chapter 11]{Cooper2019}. The
associated standard error $\sigma$ represents the statistical uncertainty of the
estimate. Under the assumption of (asymptotic) normality, confidence intervals
for $\theta$ and $p$-values for testing the null hypothesis
$H_{0} \colon \theta = 0$ can be computed as shown in Table~\ref{tab:meta}.

@@

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion is listed first and defines replication success by simultaneous
statistical significance of original and replication study along with their
effect estimates showing the same direction. This approach is also called the
``two-trials rule'' \citep{Senn2008} in drug regulation or ``vote-counting'' in
meta-analysis \citep{Cooper2019}. The criterion can similarly be defined through
one-sided $p$-values, so that the original and replication effect estimate are
not required to show the same direction as this is taken into account by the
$p$-values. Some replication projects \citep{Opensc2015, Errington2021} also
defined replication success via simultaneous non-significance of original and
replication study ($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this
definition ``replication success'' can almost always be achieved by conducting
original and replication study with very few samples so that the $p$-values are
large. The approach is also logically questionable as it could be seen as an
instance of the ``absence of evidence fallacy'' \citep{Altman1995} meaning the
failure to find evidence against the null hypothesis is erroneously interpreted
as evidence for the null hypothesis. Meta-analytic extensions of the
significance approach define replication success by significance of a combined
effect estimate.\footnote{The meta-analytic approach can also be given a
  Bayesian interpretation: A flat prior distribution for the underlying effect
  size $\theta$ is updated by the data from original and replication study.
  Replication success defined by a meta-analytic $p$-value being smaller than
  $\alpha$ is then equivalent to replication success via a Bayesian posterior
  probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
  respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$
  depending on the direction of combined estimate.} Typically, the assumption of
a common underlying effect size is seen as reasonable so that fixed-effects
meta-analysis is used for pooling. Random-effects meta-analysis has primarily
been used if more than one replication study of the same original study are
conducted since replicators are often interested in understanding
between-replication heterogeneity \citep[\eg{} in][]{Klein2018}.

The remaining criteria in Table~\ref{tab:rsdef} put more emphasis on
compatibility in effect size between the original and the replication study. For
example, the relative effect estimate $d = \that_{r}/\that_{o}$ quantifies how
much the replication effect estimate changed compared to the original, and the
smaller $d$ is the smaller the degree of replication success. Some projects also
report a confidence interval for $d$ \citep{Camerer2016, Camerer2018}, while
others ignore its uncertainty and make a binary cut at $d_{\text{min}} = 1$ to
define replication success \citep{Errington2021}. The criteria based on
confidence and prediction intervals define effect size compatibility on an
absolute scale. However, the criterion based on the original confidence interval
ignores the uncertainty from the replication, whereas the criterion based on the
replication confidence interval ignores the uncertainty from the original study.
The prediction interval criterion takes into account both sources of uncertainty
\citep{Patil2016}. Yet, declaring ``replication success'' based on the
prediction interval may be logically questionable due to its connection to the
meta-analytic $Q$-test.
% as it may also be seen as an instance of the ``absence of evidence fallacy''
% \citep{Altman1995};
That is, if the $p$-value from the $Q$-test is larger than $\alpha$ this is
equivalent to the replication effect estimate $\that_{r}$ being contained in its
$(1- \alpha)$ prediction interval. The null hypothesis of this test is defined
that the underlying effect sizes of original and replication study are the same
($H_{0} \colon \theta_{o} = \theta_{r}$), so a rejection of this null hypothesis
corresponds to demonstrating replication failure and not replication success.
Interpreting a failure to reject the null hypothesis as evidence for it is again
an instance of the ``absence of evidence fallacy''
\citep{Hedges2019}. % To establish replication success a
% test based on the null hypothesis $H_{0} \colon \theta_{o} \neq \theta_{r}$
% would be required \citep{Hedges2019}.
As in the case of defining replication success by simultaneous non-significance
of original and replication study, the mismatch of the null hypothesis of the
prediction interval criterion results in the undesirable property that
``replication success'' can almost always be achieved if the sample size of the
studies is small enough as the prediction interval becomes wider with larger
standard errors.

@@

\subsubsection{Example: Reproducibility Project Cancer Biology}
I will now illustrate the assessment of replicability on data from the
Reproducibility Project: Cancer Biology \citep{Errington2021}. This large-scale
project attempted to replicate 53 landmark studies from the field of cancer
biology. However, due to various difficulties (\eg{} missing information from
the original studies or problems in conducting the experiments) only 23 of the
53 studies could be repeated. \citet{Errington2021} report that these
experiments led to data on 158 quantitative effects. However, from the data
which they provide only 132 quantitative effects are provided with original and
replication standardized mean difference effect estimates along with standard
errors, and only they will be used in the subsequent analyses.

\begin{figure}[!htb]
<< "RPCB-graph", fig.height = 3.25 >>=
ciFormat. <- function(lower, upper) {
    paste0("[", format(round(lower, 2), nsmall = 2),
           ", ", format(round(upper, 2), nsmall = 2), "]")
}
ciFormat <- Vectorize(FUN = ciFormat.)
pFormat. <- function(p) {
    if (is.na(p)) {
        return(NA)
    } else if (p > 0.0001) {
        return(as.character(signif(p, 2)))
    } else {
        return("< 0.0001")
    }
}
pFormat <- Vectorize(FUN = pFormat.)
za <- qnorm(p = 0.975)
rpcb <- read.csv("data/rpcb.csv") |>
    mutate(zo = smdo/so,
           zr = smdr/sr,
           c = so^2/sr^2,
           d = smdr/smdo,
           po2 = 2*(1 - pnorm(q = abs(zo))),
           pr2 = 2*(1 - pnorm(q = abs(zr))),
           sm = 1/sqrt(1/so^2 + 1/sr^2),
           smdm = (smdo/so^2 + smdr/sr^2)*sm^2,
           pm2 = 2*(1 - pnorm(q = abs(smdm/sm))),
           Q = (smdo - smdr)^2/(so^2 + sr^2),
           pQ = pchisq(q = Q, df = 1, lower.tail = FALSE),
           dFormat = ifelse(d > 1,
                            paste0("\\textcolor{nicegreen}{", round(d, 2),
                                   "}"),
                            paste0("\\textcolor{red}{", round(d, 2), "}")),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,
           id = paste0("(", paper, ", ", experiment, ", ", effect, ")"),
           po2Format = ifelse(po2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(po2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(po2), "}")),
           pr2Format = ifelse(pr2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(pr2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(pr2), "}")),
           pm2Format = ifelse(pm2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(pm2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(pm2), "}")),
           pQFormat = ifelse(pQ > 0.05, paste0("\\textcolor{nicegreen}{", pFormat(pQ),
                                               "}"),
                             paste0("\\textcolor{red}{", pFormat(pQ), "}")),
           smdoFormat = paste(format(round(smdo, 2), nsmall = 2),
                              ciFormat(lower = cioL, upper = cioU)),
           smdrFormat = paste(format(round(smdr, 2), nsmall = 2),
                              ciFormat(lower = cirL, upper = cirU)),
           smdmFormat = paste(format(round(smdm, 2), nsmall = 2),
                              ciFormat(lower = cimL, upper = cimU)),
           rsig = factor(pr2 <= 0.05, levels = c(FALSE, TRUE),
                         labels = c("italic(p[r]) > 0.05",
                                    "italic(p[r]) <= 0.05")))
## compute sceptical p-value (TODO fix problems with NA)
ind <- !is.na(rpcb$zo) & !is.na(rpcb$zr) & !is.na(rpcb$c)
rpcb$ps2 <- NA
rpcb$ps2[ind] <- pSceptical(zo = rpcb$zo[ind], zr = rpcb$zr[ind], c = rpcb$c[ind],
                            alternative = "two.sided", type = "nominal")
rpcb$ps2Format <- NA
rpcb$ps2Format[ind] <- ifelse(rpcb$ps2[ind] < 0.05, paste0("\\textcolor{nicegreen}{",
                                                           pFormat(rpcb$ps2[ind]),
                                                           "}"),
                              paste0("\\textcolor{red}{", pFormat(rpcb$ps2[ind]), "}"))

@@

A detailed view for a subset of the data from the project is given in
Table~\ref{tab:RPCB}. For no single study pair in the table are all commonly
used criteria satisfied simultaneously. For instance, the two pairs (48, 2, 1)
and (16, 3, 3) at the top of the table satisfy the significance criterion
($p_{o} < 0.05$ and $p_{r} < 0.05$), meta analytic significance ($p_{m} < 0.05$)
and the $Q$-test/prediction interval criterion ($p_{Q} > 0.05$), yet the
replication effect estimate is smaller than the original one ($d < 1$).
Similarly, there is no single study pair for which all criteria indicate
replication failure. For example, the pair (5, 1, 3) at the bottom of the table
is a clear failure with respect to the relative effect estimate ($d < 1$) and
the significance criteria ($p_{o} > 0.05$ and $p_{r} > 0.05$), yet the $Q$-test
does not indicate evidence for inconsistency of the two effect estimates
($p_{Q} > 0.05$).

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the Reproducibility Project: Cancer Biology
    \citep{Errington2021}. (P, X, E) denotes effect number E from experiment
    number X, from original paper number P. Shown are original $\that_{o}$,
    replication $\that_{r}$, and pooled effect estimate $\that_{m}$ with 95\%
    confidence intervals, variance ratio $c = \sigma^{2}_{o}/\sigma^{2}_{r}$,
    relative effect estimate $d = \that_{r}/\that_{o}$, original
    \textit{p}-value $p_{o}$, replication \textit{p}-value $p_{r}$,
    meta-analytic \textit{p}-value $p_{m}$, $Q$-test \textit{p}-value $p_{Q}$,
    and sceptical \textit{p}-value $p_{\mathrm{S}}$. A level $\alpha = 5\%$ is
    used as threshold for replication success via \textit{p}-values, and a
    threshold $d_{\text{min}} = 1$ for the relative effect estimate.}
  \label{tab:RPCB}
% \resizebox{1\textwidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    filter(smdo < 80, smdr < 80) |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(pr2) |>
    select(id, smdoFormat, ## so,
           smdrFormat, smdmFormat,
           c, dFormat, ## sr,
           po2Format, pr2Format, pm2Format, pQFormat, ps2Format
           )

rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$ [95\\% CI]",
                          ## "$\\sigma_o$",
                          "$\\that_r$ [95\\% CI]",
                          "$\\that_m$ [95\\% CI]",
                          "$c$", #"$\\sigma^2_o/\\sigma^2_r$",
                          "$d$", #"$\\that_r/\\that_o$",
                          ## "$\\sigma_r$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )
print(rpcbXtable, booktabs = TRUE, floating = FALSE, include.rownames = FALSE,
      sanitize.text.function = function(x) {x},
      size = "small")
@
% %
% }
\end{table}

@@

To illustrate reverse-Bayes inference, let us return to the replication setting.
Assume we want to conduct inference about the unknown effect size $\theta$ based
on the effect estimate from the original study $\that_{o}$ and its standard
error $\sigma_{o}$. We will assume that $\that_{o}$ is normally distributed
around the unknown effect size $\theta$ with (known) variance equal to its
squared standard error $\sigma^{2}_{o}$, here and henceforth denoted by
$\that_{o} \given \theta \sim \Nor(\theta, \sigma^{2}_{o})$. Furthermore, we
specify a zero-mean normal prior with variance $\tau^{2}$ for the effect size
$\theta$, representing the position of a sceptic who does not believe in the
presence of a genuine effect. The ``stubbornness'' of the sceptic is determined
by how large the variance $\tau^{2}$ is chosen. Combining the sceptical prior
with the likelihood produces a posterior which is again normal
$\theta \given \that_{o}, \sigma_{o} \sim \Nor(\mu_{\text{post}}, \sigma^{2}_{\text{post}})$
with mean and variance
\begin{align*}
  &\mu_{\text{post}} = \frac{\that_{o}}{1 + \sigma^{2}_{o}/\tau^{2}}&
  &\text{and}&
  &\sigma^{2}_{\text{post}} = \frac{1}{1/\sigma^{2}_{o} + 1/\tau^{2}}.&
\end{align*}
The associated $(1- \alpha)$ highest posterior density credible interval is
given by
\begin{align}
  \left[\mu_{\text{post}} \pm z_{\alpha/2} \, \sigma_{\text{post}}\right]
  \label{eq:postCI}
\end{align}
and if this credible interval excludes parameter values smaller/larger than zero
(depending on the orientation of the effect size) this may be interpreted as
evidence\footnote{For readers who do not agree with this notion of evidence, do
  not panic. In this thesis (Paper I and IV) we extend this reverse-Bayes
  procedure to use alternative measures of evidence, such as Bayes factors.} for
a genuine effect found in the original study.

@@

The sceptical $p$-value has several interesting properties \citep[section
3]{Held2020}: First, it is always larger than the maximum of the original and
replication $p$-values ($p_{\text{S}} > \max\{p_{o}, p_{r}\}$), meaning that
both $p$-values have to be smaller than $\alpha$ such that replication success
based on $p_{\text{S}} \leq \alpha$ is possible. Second, if the $p$-values
$p_{o}$ and $p_{r}$ remain fixed but the relative effect estimate
$d = \that_{r}/\that_{o}$ decreases, the sceptical $p$-value increases, meaning
that shrinkage of the replication effect estimate is penalized
($p_{\text{S}} \uparrow 1$ as $d \downarrow 0$ for fixed $p_{o}$ and $p_{r}$).
Because of the first property, the sceptical $p$-value requires \emph{both}
studies to be sufficiently convincing on their own (in terms of their
$p$-values), similar to the significance criterion for replication success.
However, unlike the significance criterion, if the $p$-values remain fixed but
the replication effect estimate $\that_{r}$ becomes smaller than the original
estimate $\that_{o}$, the sceptical $p$-value becomes larger so indicates less
replication success due to the second property. This is desirable in the
replication setting as a smaller replication effect estimate $\that_{r}$ may not
be practically relevant anymore, despite its statistical significance.

>>>>

\chapter*{Introduction}
\phantomsection
\addcontentsline{toc}{chapter}{\texorpdfstring{\phantom{IV}Introduction}{Introduction}}

@@

How can we know if a finding from a scientific study is really true? For
example, how can we know if the protective effect of a vaccine found in a study
is real? The answer to such questions is of high importance to scientists,
decision makers, and the general population. However, because any study result
comes with uncertainty we can never know for sure.

@@

Despite their tremendous importance, the traditional academic system has made it
unattractive for researchers to conduct replication studies; the currencies of
science -- publications, citations, and grant money -- are typically easier to
acquire by conducting novel research. This is because until relatively recently,
top journals in many fields refused to publish replication studies as they were
not considered ``innovative'' enough. To build a successful career, researchers
thus often had no other choice than to concentrate their efforts on producing
novel and eye-catching research results \citep{Martin2017}.

@@

\subsection{Statistical aspects of replication studies}
Despite the increased interest in replication studies, the research community
has not yet agreed on one important question: When is a replication study
considered successful? For this reason, replication researchers typically report
the results from different methods for assessing replication success. For
example, the \citet{Opensc2015} states ``[t]here is no single standard for
evaluating replication success. We evaluated [replicability] using significance
and P values, effect sizes, subjective assessments of replication teams, and
meta-analyses of effect sizes'' (p. 11). In the following, I will give an
overview about these and other methods which have been used in practice.

@@

Most methods for analyzing replication studies can be formulated in terms of
study-level summary statistics as shown in Table~\ref{tab:meta}. % Effect
% estimates, standard error, confidence intervals, and $p$-values
All of these are routinely reported in researcher articles, and if one of them
is missing they typically can, under some assumptions, be calculated from the
others. Using summary statistics is also often the only possible way for
conducting the replication success analysis as the raw data from the original
study may not be available to the replicators. The most important statistic is
the effect estimate $\that$. It provides an estimate of the underlying effect
size $\theta$ which quantifies the true effect or association of an
intervention/exposure with an outcome variable. Typical effect sizes are mean
differences and correlations (for continuous outcomes), odds ratios (for binary
outcomes), or hazard ratios (for time to event outcomes). Depending on the type
of effect size, a transformation might be required so that the assumption of
approximately normally distributed effect estimates around the unknown effect
size $\theta$ (for large enough sample sizes) is justified. This could be, for
instance, the Fisher $z$-transformation for correlations or the
log-transformation for odds/hazard ratios \citep[chapter 11]{Cooper2019}. The
associated standard error $\sigma$ represents the statistical uncertainty of the
estimate. Under the assumption of (asymptotic) normality, confidence intervals
for $\theta$ and $p$-values for testing the null hypothesis
$H_{0} \colon \theta = 0$ can be computed as shown in Table~\ref{tab:meta}.

@@

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion is listed first and defines replication success by simultaneous
statistical significance of original and replication study along with their
effect estimates showing the same direction. This approach is also called the
``two-trials rule'' \citep{Senn2008} in drug regulation or ``vote-counting'' in
meta-analysis \citep{Cooper2019}. The criterion can similarly be defined through
one-sided $p$-values, so that the original and replication effect estimate are
not required to show the same direction as this is taken into account by the
$p$-values. Some replication projects \citep{Opensc2015, Errington2021} also
defined replication success via simultaneous non-significance of original and
replication study ($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this
definition ``replication success'' can almost always be achieved by conducting
original and replication study with very few samples so that the $p$-values are
large. The approach is also logically questionable as it could be seen as an
instance of the ``absence of evidence fallacy'' \citep{Altman1995} meaning the
failure to find evidence against the null hypothesis is erroneously interpreted
as evidence for the null hypothesis. Meta-analytic extensions of the
significance approach define replication success by significance of a combined
effect estimate.\footnote{The meta-analytic approach can also be given a
  Bayesian interpretation: A flat prior distribution for the underlying effect
  size $\theta$ is updated by the data from original and replication study.
  Replication success defined by a meta-analytic $p$-value being smaller than
  $\alpha$ is then equivalent to replication success via a Bayesian posterior
  probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
  respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$
  depending on the direction of combined estimate.} Typically, the assumption of
a common underlying effect size is seen as reasonable so that fixed-effect
meta-analysis is used for pooling. Random-effects meta-analysis has primarily
been used if more than one replication study of the same original study are
conducted since replicators are often interested in understanding
between-replication heterogeneity \citep[\eg{} in][]{Klein2018}.

The remaining criteria in Table~\ref{tab:rsdef} put more emphasis on
compatibility in effect size between the original and the replication study. For
example, the relative effect estimate $d = \that_{r}/\that_{o}$ quantifies how
much the magnitude of the replication effect estimate changed compared to the
original, and the smaller $d$ is the smaller the degree of replication success.
Some projects also report a confidence interval for $d$ \citep{Camerer2016,
  Camerer2018}, while others ignore its uncertainty and make a binary cut at
$d_{\text{min}} = 1$ to define replication success \citep{Errington2021}. The
criteria based on confidence and prediction intervals define effect size
compatibility on an absolute scale. However, the criterion based on the original
confidence interval ignores the uncertainty from the replication, whereas the
criterion based on the replication confidence interval ignores the uncertainty
from the original study. The prediction interval criterion takes into account
both sources of uncertainty \citep{Patil2016}. Yet, declaring ``replication
success'' based on the prediction interval may be logically questionable due to
its connection to the meta-analytic $Q$-test.
% as it may also be seen as an instance of the ``absence of evidence fallacy''
% \citep{Altman1995};
That is, if the $p$-value from the $Q$-test is larger than $\alpha$ this is
equivalent to the replication effect estimate $\that_{r}$ being contained in its
$(1- \alpha)$ prediction interval. The null hypothesis of this test is defined
that the underlying effect sizes of original and replication study are the same
($H_{0} \colon \theta_{o} = \theta_{r}$), so a rejection of this null hypothesis
corresponds to demonstrating replication failure and not replication success.
Interpreting a failure to reject the null hypothesis as evidence for it is again
an instance of the ``absence of evidence fallacy''
\citep{Hedges2019}. % To establish replication success a
% test based on the null hypothesis $H_{0} \colon \theta_{o} \neq \theta_{r}$
% would be required \citep{Hedges2019}.
As in the case of defining replication success by simultaneous non-significance
of original and replication study, the mismatch of the null hypothesis of the
prediction interval criterion results in the undesirable property that
``replication success'' can almost always be achieved if the sample size of the
studies is small enough as the prediction interval becomes wider with larger
standard errors.

@@

\subsubsection{Example: Reproducibility Project Cancer Biology}
I will now illustrate the assessment of replicability on data from the
Reproducibility Project: Cancer Biology \citep{Errington2021}. This large-scale
project attempted to replicate 53 landmark studies from the field of cancer
biology but due to various difficulties (\eg{} missing information from the
original studies or problems in conducting the experiments) only 23 of the 53
studies could be repeated. \citet{Errington2021} report that these experiments
led to data on 158 quantitative effects. However, from the data which they
provide only 132 quantitative effects are provided with original and replication
standardized mean difference effect estimates along with standard errors, and
only they will be used in the subsequent analyses.

\begin{figure}[!htb]
<< "RPCB-graph", fig.height = 3.25 >>=
ciFormat. <- function(lower, upper) {
    paste0("[", format(round(lower, 2), nsmall = 2),
           ", ", format(round(upper, 2), nsmall = 2), "]")
}
ciFormat <- Vectorize(FUN = ciFormat.)
pFormat. <- function(p) {
    if (is.na(p)) {
        return(NA)
    } else if (p > 0.0001) {
        return(as.character(signif(p, 2)))
    } else {
        return("< 0.0001")
    }
}
pFormat <- Vectorize(FUN = pFormat.)
za <- qnorm(p = 0.975)
rpcb <- read.csv("data/rpcb.csv") |>
    mutate(zo = smdo/so,
           zr = smdr/sr,
           c = so^2/sr^2,
           d = smdr/smdo,
           po2 = 2*(1 - pnorm(q = abs(zo))),
           pr2 = 2*(1 - pnorm(q = abs(zr))),
           sm = 1/sqrt(1/so^2 + 1/sr^2),
           smdm = (smdo/so^2 + smdr/sr^2)*sm^2,
           pm2 = 2*(1 - pnorm(q = abs(smdm/sm))),
           Q = (smdo - smdr)^2/(so^2 + sr^2),
           pQ = pchisq(q = Q, df = 1, lower.tail = FALSE),
           dFormat = ifelse(d > 1,
                            paste0("\\textcolor{nicegreen}{$", round(d, 2),
                                   "$}"),
                            paste0("\\textcolor{red}{$", round(d, 2), "$}")),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,
           id = paste0("(", paper, ", ", experiment, ", ", effect, ")"),
           po2Format = ifelse(po2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(po2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(po2), "}")),
           pr2Format = ifelse(pr2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(pr2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(pr2), "}")),
           pm2Format = ifelse(pm2 < 0.05,
                              paste0("\\textcolor{nicegreen}{", pFormat(pm2),
                                     "}"),
                              paste0("\\textcolor{red}{", pFormat(pm2), "}")),
           pQFormat = ifelse(pQ > 0.05, paste0("\\textcolor{nicegreen}{", pFormat(pQ),
                                               "}"),
                             paste0("\\textcolor{red}{", pFormat(pQ), "}")),
           smdoFormat = paste("$", format(round(smdo, 2), nsmall = 2),
                              ciFormat(lower = cioL, upper = cioU), "$"),
           smdrFormat = paste("$", format(round(smdr, 2), nsmall = 2),
                              ciFormat(lower = cirL, upper = cirU), "$"),
           smdmFormat = paste("$", format(round(smdm, 2), nsmall = 2),
                              ciFormat(lower = cimL, upper = cimU), "$"),
           rsig = factor(pr2 <= 0.05, levels = c(FALSE, TRUE),
                         labels = c("italic(p[r]) > 0.05",
                                    "italic(p[r]) <= 0.05")))
## compute sceptical p-value (TODO fix problems with NA)
ind <- !is.na(rpcb$zo) & !is.na(rpcb$zr) & !is.na(rpcb$c)
rpcb$ps2 <- NA
rpcb$ps2[ind] <- pSceptical(zo = rpcb$zo[ind], zr = rpcb$zr[ind], c = rpcb$c[ind],
                            alternative = "two.sided", type = "nominal")
rpcb$ps2Format <- NA
rpcb$ps2Format[ind] <- ifelse(rpcb$ps2[ind] < 0.05, paste0("\\textcolor{nicegreen}{",
                                                           pFormat(rpcb$ps2[ind]),
                                                           "}"),
                              paste0("\\textcolor{red}{", pFormat(rpcb$ps2[ind]), "}"))

@@

A detailed view for a subset of the data from the project is given in
Table~\ref{tab:RPCB}. The subset is chosen such that from every original study
one effect and one replication of that effect is included. For no single study
pair in the table are all commonly used criteria satisfied simultaneously. For
instance, the first two replications satisfy the significance criterion
($p_{o} < 0.05$ and $p_{r} < 0.05$), meta analytic significance ($p_{m} < 0.05$)
and the $Q$-test/prediction interval criterion ($p_{Q} > 0.05$), yet the
replication effect estimate is smaller than the original one ($d < 1$).
Similarly, there is no single study pair for which all criteria indicate
replication failure. For example, the last replication (\#20) is a clear failure
with respect to the relative effect estimate ($d < 1$) and the significance
criteria ($p_{o} > 0.05$ and $p_{r} > 0.05$), yet the $Q$-test does not indicate
evidence for inconsistency of the two effect estimates ($p_{Q} > 0.05$).

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the Reproducibility Project: Cancer Biology
    \citep{Errington2021}. (P, X, E) denotes effect number E from experiment
    number X, from original paper number P. Shown are original $\that_{o}$,
    replication $\that_{r}$, and pooled effect estimate $\that_{m}$ with 95\%
    confidence intervals, variance ratio $c = \sigma^{2}_{o}/\sigma^{2}_{r}$,
    relative effect estimate $d = \that_{r}/\that_{o}$, original
    \textit{p}-value $p_{o}$, replication \textit{p}-value $p_{r}$,
    meta-analytic \textit{p}-value $p_{m}$, $Q$-test \textit{p}-value $p_{Q}$,
    and sceptical \textit{p}-value $p_{\mathrm{S}}$. A level $\alpha = 5\%$ is
    used as threshold for replication success via \textit{p}-values, and a
    threshold $d_{\text{min}} = 1$ for the relative effect estimate.}
  \label{tab:RPCB}
% \resizebox{1\textwidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    ## filter(smdo < 80, smdr < 80) |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(pr2) |>
    select(id,
           smdoFormat, ## so,
           smdrFormat, smdmFormat,
           c, dFormat, ## sr,
           po2Format, pr2Format, pm2Format, pQFormat, ps2Format
           )

rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$ [95\\% CI]",
                          ## "$\\sigma_o$",
                          "$\\that_r$ [95\\% CI]",
                          "$\\that_m$ [95\\% CI]",
                          "$c$", #"$\\sigma^2_o/\\sigma^2_r$",
                          "$d$", #"$\\that_r/\\that_o$",
                          ## "$\\sigma_r$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )
print(rpcbXtable, booktabs = TRUE, floating = FALSE, include.rownames = TRUE,
      sanitize.text.function = function(x) {x},
      size = "small")
@
% %
% }
\end{table}

@@

To illustrate reverse-Bayes inference, let us return to the replication setting.
Assume we want to conduct inference about the unknown effect size $\theta$ based
on the effect estimate from the original study $\that_{o}$ and its standard
error $\sigma_{o}$. We will assume that $\that_{o}$ is normally distributed
around the unknown effect size $\theta$ with (known) variance equal to its
squared standard error $\sigma^{2}_{o}$, here and henceforth denoted by
$\that_{o} \given \theta \sim \Nor(\theta, \sigma^{2}_{o})$. Furthermore, we
specify a zero-mean normal prior with variance $\tau^{2}$ for the effect size
$\theta$, representing the position of a sceptic who does not believe in the
presence of a genuine effect. The ``stubbornness'' of the sceptic is determined
by how large the variance $\tau^{2}$ is chosen. Combining the sceptical prior
with the likelihood produces a posterior which is again normal
$\theta \given \that_{o}, \sigma_{o} \sim \Nor(\mu_{\text{post}}, \sigma^{2}_{\text{post}})$
with mean and variance
\begin{align*}
  &\mu_{\text{post}} = \frac{\that_{o}}{1 + \sigma^{2}_{o}/\tau^{2}}&
  &\text{and}&
  &\sigma^{2}_{\text{post}} = \frac{1}{1/\sigma^{2}_{o} + 1/\tau^{2}}.&
\end{align*}
The associated $(1- \alpha)$ highest posterior density credible interval is
given by
\begin{align}
  \left[\mu_{\text{post}} \pm z_{\alpha/2} \, \sigma_{\text{post}}\right]
  \label{eq:postCI}
\end{align}
and if this credible interval excludes parameter values smaller/larger than zero
(depending on the orientation of the effect size) this may be interpreted as
evidence\footnote{For readers who do not agree with this notion of evidence, do
  not panic. In this thesis (Paper II and IV) we extend this reverse-Bayes
  procedure to use alternative measures of evidence, such as Bayes factors.} for
a genuine effect found in the original study.

@@

The sceptical $p$-value has several interesting properties \citep[section
3]{Held2020}: First, it is always larger than the maximum of the original and
replication $p$-values ($p_{\text{S}} > \max\{p_{o}, p_{r}\}$), meaning that
both $p$-values have to be smaller than $\alpha$ such that replication success
at level $\alpha$ is possible. The sceptical $p$-value hence requires
\emph{both} studies to be sufficiently convincing on their own (in terms of
their $p$-values), similar to the significance criterion for replication
success. Second, if the $p$-values $p_{o}$ and $p_{r}$ remain fixed but the
relative effect estimate $d = \that_{r}/\that_{o}$ decreases, the sceptical
$p$-value increases, meaning that shrinkage of the replication effect estimate
is penalized ($p_{\text{S}} \uparrow 1$ as $d \downarrow 0$ for fixed $p_{o}$
and $p_{r}$). The sceptical $p$-value thus takes effect shrinkage into account,
unlike the significance criterion which can be achieved with any non-zero
replication effect estimate $\that_{r}$ provided the standard error $\sigma_{r}$
is small enough. This property is desirable in the replication setting as a
smaller replication effect estimate $\that_{r}$ may not be practically relevant
anymore, despite its statistical significance.
